Viscous Transport in Eroding Porous Media

### 2019-09-11 docs/transport/erosion_transport.tex
working on edits to intro

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
\label{sec:intro}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Flow in many natural occurring geometries involves flows that vary over
many orders of magnitude.  \todo[]{The grammar in the first sentence is awkward: Flow in ... involves flows?}
For example, the length scales of a porous
media can vary from $10^{-6}m$ to
$10^{-1}m$~\cite{mil-chr-imh-mcb-ped1998}.  Porous media flow plays an
important role in several environmental and industrial applications.
Therefore, understanding the dynamics of pore-scale flow is central to
upscale the governing fluid equations, characterize
dispersion~\cite{saf1959}, and quantify mixing. However, because of the
large disparity of length and velocity scales, constitutive laws that
link the microscopic scale to macroscopic scale must be
developed~\cite{mil-chr-imh-mcb-ped1998}.  Therefore, to find closure
formulas, accurate simulations at microscopic scales must be performed
and analyzed.  A few examples of closure formulas and coarse-grained
variables include
pressure-saturation-permeability~\cite{mil-chr-imh-mcb-ped1998},
permeability-porosity~\cite{dar-mcc1998, car1937, won-kop-tom1984},
first passage time distribution~\cite{ber-sch-sil2000,
hym-den-hag-kan2019, cve-che-wen1996}, tortuosity~\cite{hak-com-den2019,
mat-kha-koz2008, dud-koz-mat2011, kop-kat-tim1996}, concentration
distributions~\cite{ica-den2019, bel-sal-rin1992}, flow
distributions~\cite{ali-par-wei-bre2017}, geometry
connectivity~\cite{knu-car2005, wes-blo-gra2001}, pore
distributions~\cite{ali-par-wei-bre2017}, and anomalous
dispersion~\cite{dea-qua-bir-jua2018, den-cor-sch-ber2004,
sie-ili-pri-riv-gua2019}.

In many applications, the fluid transport continuously changes the pore
structure.  This is particularly common in geophysical flows where
erosion alters the porous matrix.  In this work, we investigate flow in
eroded porous media.  By introducing erosion, there is a two-way
coupling between the hydrodynamics and the mechanical alteration of the
complex porous geometry.  Transport in porous geometries that have
undergone erosion finds applications in aquifers~\cite{cve-che-wen1996,
sud1986}, solute transport in groundwater flow~\cite{dag1987,
kon-bre1978}.  Porous geometries that have undergone erosion have
certain characteristics.  For example, the geometry often has regions of
high porosity that connect the inlet to the outlet.  This channelization
results in velocity scales that can vary over several orders of
magnitude~\cite{all-hea-lab-rei2002}.  Moreover, the channelization
results in a geometry that is largely heterogeneous and anisotropic, and
this affects heat transfer through the medium~\cite{nil-sto1990,
ree-sto1995}.  Others have studied transport in heterogeneous geometries
(for example, see~\cite{mil-chr-imh-mcb-ped1998, ber-sch-sil2000,
hak-com-den2019, suo-liu-gan2019, cus-hu-den1995, cve-che-wen1996,
leb-ded-dav-bou2007, knu-car2005, bel-sal-rin1992}). However, to the
best of our knowledge, a detailed study of transport in eroded
geometries has not been performed.

To describe the relevant hydrodynamics, we model two-dimensional
microscopic scales where the fluid is governed by the incompressible
Stokes equations (for example, see~\cite{hym-den-hag-kan2019,
den-ica-hid2018, bar-mar-vee-zha2018}).  To couple the erosion and the
hydrodynamics, the grains of the porous media are eroded at a rate that
is proportional to the hydrodynamic shear stress~\cite{wan-fel2004,
ris-moo-chi-she-zha2012, qua-moo2018, par-izu2000}.  A variety of
numerical methods can be applied to solve the incompressible Stokes
equations such as a lattice-gas or lattice Boltzmann
model~\cite{kop-kat-tim1996, dar-mcc1998}, finite
volume~\cite{suo-liu-gan2019, den-ica-hid2018, sie-ili-pri-riv-gua2019},
finite differences~\cite{leb-ded-dav-bou2007, knu-car2005}, and cellular
automata~\cite{rot1988}. In this work, we instead solve an equivalent
boundary integral equation (BIE) formulation of the incompressible
Stokes equation.  The BIE is discretized with a high-order quadrature
method that resolves nearly touching bodies with a near-singular
integration scheme~\cite{bar-wu-vee2015}.  The erosion dynamics are
simulated by applying a stable second-order time stepping Runge-Kutta
method to the {\thL} formulation~\cite{hou-low-she1994} of the eroding
bodies.  Given the solution of the incompressible Stokes equations are
solved in an eroded porous geometry, we investigate different metrics to
characterize transport.  We focus on three metrics: the tortuosity, the
anomalous dispersion, and the distribution of geometrical channels.

The first two of these metrics can be defined in terms of statistical
moments of trajectories.  Depending on the application, transport at the
microscale can be modelled as pure advection~\cite{dea-qua-bir-jua2018,
leb-ded-dav-bou2007, cve-che-wen1996, puy-gou-den2019}, and advection
dispersion equation~\cite{cus-hu-den1995, dag1987, den-ica-hid2018}, or
with a random walk~\cite{saf1959, bij-blu2006, ber-sch-sil2000}.  In
this work, we assume that the trajectories are governed purely by
advection (ie.~no diffusion).  Therefore, a trajectory $\ss(t)$
initialized at $\ss_0$ is governed by
\begin{align}
  \frac{d\ss}{dt} = \uu(\ss,t), \quad \ss(0) = \ss_0,
  \label{eqn:tracers}
\end{align}
where $\uu$ is the fluid velocity.  Since there is no diffusion, all
dispersion is a result of the particle spreading induced by the porous
geometry.  Since there is no diffusion, we apply a high-order explicit
Runge-Kutta time stepping method to compute streamlines.  However, other
groups have simulated trajectories using continuous time random
walks~\cite{mon-wei1965, den-cor-sch-ber2004, leb-den-car2008,
ber-cor-den-sch2006} and a Fokker-Planck equation~\cite{ica-den2019}.

\todo[inline]{Dump of text}
Resolving the flow in these small velocity regions is essential to
accurately describe anomalous dispersion and
mixing~\cite{leb-den-dav-bol-car-dec-bou2011, den-leb-eng-bij2011}.
\todo[inline]{Dump of text}

@@

The numerical methods to solve the governing fluid and erosion equations
are based on previous work of the authors~\cite{qua-moo2018}.  We
continue to use a boundary integral equation (BIE) formulation of the
incompressible Stokes equations and a second-order Runge-Kutta method
applied to a smoothed version of the magnitude of the shear stress.  In
our previous work, we discretized the BIE with the trapezoid rule whose
accuracy grows when evaluating layer potentials close to the fluid
boundary.  Therefore, given a fixed resolution, our previous work
required a minimum separation distance between all bodies to maintain
stability.  One of the earliest works on near-singular integration was
developed by Baker and Shelley~\cite{bak-she1986}, and in recent years,
many other near-singular integration schemes have been
developed~\cite{oja-tor2015, kli-tor2018, hel-oja2008a, bea-yin-wil2016,
bea-lai2001, klo-bar-gre-one2013}.  In this work, we use a Barycentric
quadrature method~\cite{bar2014, bar-wu-vee2015} since it is a
non-intrusive modification of the trapezoid rule, and the error is
guaranteed to be uniformly bounded.  In this work, we extend this
quadrature method so that it can be used to compute the gradient of the
velocity which is required to compute the shear stress and the fluid
vorticity. 
\begin{figure}[H]
\begin{center}
\includegraphics[width = 0.3 \textwidth]{./figs/vort_50b1}
\includegraphics[width = 0.3 \textwidth]{./figs/vort_50b80}
\includegraphics[width = 0.38 \textwidth]{./figs/vort_50b160}\\
\includegraphics[width = 0.3 \textwidth]{./figs/vort_50b240}
\includegraphics[width = 0.3 \textwidth]{./figs/vort_50b320}
\includegraphics[width = 0.38 \textwidth]{./figs/vort_50b400}
\caption{\label{fig:Eroding50vort} Erosion of 50 nearly touching
bodies with the fixed pressure drop condition. The color is the
vorticity of the fluid. The 6 snapshots are evenly spaced in time.}
\end{center}
\end{figure}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
\label{sec:intro}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Porous media flow plays an important role in several environmental and
industrial applications.  To understand flow in a porous media, it is
important to accurately characterize the length scales and velocity
scales. However, this is very challenging since for a single porous
geometry, the length scales that can vary from $10^{-6}m$ to
$10^{-1}m$~\cite{mil-chr-imh-mcb-ped1998}, and the velocity scales that
can vary from $10^{-1}m/$day to $10m/$day~\cite{kut-scr-dav-ham1995}.
However, resolving all these scales is fundamental to develop meaningful
constitutive relationships that link the microscopic to the macroscopic
scale~\cite{mil-chr-imh-mcb-ped1998}, characterize
dispersion~\cite{saf1959}, and quantify
mixing~\cite{leb-den-dav-bol-car-dec-bou2011, den-leb-eng-bij2011}.
Coarse-grained variables are often used to study flow in a porous media,
and examples include pressure-saturation-permeability
relationships~\cite{mil-chr-imh-mcb-ped1998}, permeability-porosity
relationships~\cite{dar-mcc1998, car1937, won-kop-tom1984}, first
passage time distributions~\cite{ber-sch-sil2000, hym-den-hag-kan2019,
cve-che-wen1996}, tortuosities~\cite{hak-com-den2019, mat-kha-koz2008,
dud-koz-mat2011, kop-kat-tim1996}, concentration
distributions~\cite{ica-den2019, bel-sal-rin1992}, flow
distributions~\cite{ali-par-wei-bre2017}, geometry
connectivity~\cite{knu-car2005, wes-blo-gra2001}, pore
distributions~\cite{ali-par-wei-bre2017}, anomalous
dispersion~\cite{dea-qua-bir-jua2018, den-cor-sch-ber2004,
sie-ili-pri-riv-gua2019}, and more.

Flow in porous media is often studied for a fixed geometry, but for many
physical processes, the geometry evolves with the flow. For example,
melting~\cite{}\todo[inline]{Natural convection solid/liquid phase
change in porous media}, dissolution~\cite{kan-zha-che-he2002},
deposition~\cite{joh-eli1995}, and biofilm
growth~\cite{tan-val-wer2015}.  Here, we focus on erosion which is
fluid-mechanical process that is prevalent in many geophysical and
hydrological applications.  For example, contaminant transport through
aquifers~\cite{cve-che-wen1996} and groundwater~\cite{dag1987,
kon-bre1978} is largely effected by the nature of the eroded porous
geometry.  Porous geometries that have undergone erosion have
certain characteristics.  For example, the geometry often has regions of
high porosity that connect the inlet to the outlet.  This channelization
results in velocity scales that vary over several orders of
magnitude~\cite{all-hea-lab-rei2002} (see
Figure~\ref{fig:Eroding50vel}).  Moreover, the channelization results in
a geometry that is largely heterogeneous and anisotropic, and this
affects the transfer other other quantities such as a contaminant or
heat~\cite{nil-sto1990,
ree-sto1995}.  Others have studied transport in heterogeneous geometries
(for example, see~\cite{mil-chr-imh-mcb-ped1998, ber-sch-sil2000,
hak-com-den2019, suo-liu-gan2019, cus-hu-den1995, cve-che-wen1996,
leb-ded-dav-bou2007, knu-car2005, bel-sal-rin1992}). However, to the
best of our knowledge, a detailed study of transport in eroded
geometries has not been performed.

To describe the relevant hydrodynamics, we model two-dimensional
microscopic scales where the fluid is governed by the incompressible
Stokes equations (for example, see~\cite{hym-den-hag-kan2019,
den-ica-hid2018, bar-mar-vee-zha2018}).  To couple the erosion and the
fluid, the grains of the porous media are eroded at a rate that
is proportional to the hydrodynamic shear stress~\cite{wan-fel2004,
ris-moo-chi-she-zha2012, qua-moo2018, par-izu2000}.  A variety of
numerical methods can be applied to solve the incompressible Stokes
equations such as a lattice-gas or lattice Boltzmann
model~\cite{kop-kat-tim1996, dar-mcc1998}, finite
volume~\cite{suo-liu-gan2019, den-ica-hid2018, sie-ili-pri-riv-gua2019},
finite differences~\cite{leb-ded-dav-bou2007, knu-car2005}, and cellular
automata~\cite{rot1988}. In this work, we instead solve an equivalent
boundary integral equation (BIE) formulation of the incompressible
Stokes equation.  The BIE is discretized with a high-order quadrature
method that resolves nearly touching bodies with a near-singular
integration scheme~\cite{bar-wu-vee2015}.  The erosion dynamics are
simulated by applying a stable second-order time stepping Runge-Kutta
method to the {\thL} formulation~\cite{hou-low-she1994} of the eroding
bodies.  Given the solution of the incompressible Stokes equations in
an eroded porous geometry, we investigate different different but
related metrics to
characterize transport.  We focus on three metrics: the tortuosity, the
anomalous dispersion, and the distribution of geometrical channels.

The first two of these metrics can be defined in terms of statistical
moments of trajectories.  Depending on the application, transport at the
microscale can be modelled as pure advection~\cite{dea-qua-bir-jua2018,
leb-ded-dav-bou2007, cve-che-wen1996, puy-gou-den2019}, and advection
diffusion equation~\cite{cus-hu-den1995, dag1987, den-ica-hid2018}, or
with a random walk~\cite{saf1959, bij-blu2006, ber-sch-sil2000}.  In
this work, we assume that the trajectories are governed purely by
advection (ie.~no diffusion).  Therefore, a trajectory $\ss(t)$
initialized at $\ss_0$ is governed by
\begin{align}
  \frac{d\ss}{dt} = \uu(\ss,t), \quad \ss(0) = \ss_0,
  \label{eqn:tracers}
\end{align}
where $\uu$ is the fluid velocity.  Tracer trajectories can be simulated
with a continuous time random walks~\cite{mon-wei1965,
den-cor-sch-ber2004, leb-den-car2008, ber-cor-den-sch2006} or with a
Fokker-Planck equation~\cite{ica-den2019}.  Instead, since there is no
stiff diffusive term in~\eqref{eqn:tracers}, we compute streamlines by
applying a high-order explicit Runge-Kutta time stepping method.

\todo[inline]{Bryan left off here}

@@

The numerical methods to solve the governing fluid and erosion equations
are based on previous work of the authors~\cite{qua-moo2018}.  We
continue to use a boundary integral equation (BIE) formulation of the
incompressible Stokes equations and a second-order Runge-Kutta method
applied to a smoothed version of the magnitude of the shear stress.  In
our previous work, we discretized the BIE with the trapezoid rule whose
accuracy grows when evaluating layer potentials close to the fluid
boundary.  Therefore, given a fixed resolution, our previous work
required a minimum separation distance between all bodies to maintain
stability.  One of the earliest works on near-singular integration was
developed by Baker and Shelley~\cite{bak-she1986}, and in recent years,
many other near-singular integration schemes have been
developed~\cite{oja-tor2015, kli-tor2018, hel-oja2008a, bea-yin-wil2016,
bea-lai2001, klo-bar-gre-one2013}.  In this work, we use a Barycentric
quadrature method~\cite{bar2014, bar-wu-vee2015} since it is a
non-intrusive modification of the trapezoid rule, and the error is
guaranteed to be uniformly bounded.  In this work, we extend this
quadrature method so that it can be used to compute the gradient of the
velocity which is required to compute the shear stress and the fluid
vorticity.

\begin{figure}[H]
\begin{center}
\todo[inline]{plots of the magnitude of the velocity on a log based 10
scale}
\caption{\label{fig:Eroding50vel} Erosion of 50 nearly touching bodies
with the fixed pressure drop condition. The color is the vorticity of
the fluid. The 6 snapshots are evenly spaced in time.}
\end{center}
\end{figure}

\begin{figure}[H]
\begin{center}
\includegraphics[width = 0.3 \textwidth]{./figs/vort_50b1}
\includegraphics[width = 0.3 \textwidth]{./figs/vort_50b80}
\includegraphics[width = 0.38 \textwidth]{./figs/vort_50b160}\\
\includegraphics[width = 0.3 \textwidth]{./figs/vort_50b240}
\includegraphics[width = 0.3 \textwidth]{./figs/vort_50b320}
\includegraphics[width = 0.38 \textwidth]{./figs/vort_50b400}
\caption{\label{fig:Eroding50vort} Erosion of 50 nearly touching
bodies with the fixed pressure drop condition. The color is the
vorticity of the fluid. The 6 snapshots are evenly spaced in time.}
\end{center}
\end{figure}
